---
title: "Project 1: Product Sales Insights"
author: "Someshwar"
date: "`r Sys.Date()`"
output: word_document
---

# Introduction

This project explores whether environmental and calendar-related factors are associated with daily product sales for a large retailer with 45 nationwide locations. The central business question is: **how does weather impact sales of the top three best-selling products?**

The analysis focuses on the **three products with the highest total units sold**, identified via SQL queries. Weather features — including average temperature, precipitation, and weather event flags — are used to model sales behaviour alongside calendar features such as weekend timing.

Two interpretable modelling techniques are applied per product: **linear regression** and **decision tree regression**, enabling both numerical coefficient interpretation and visual rule-based insights.

The goal is exploratory rather than purely predictive. The emphasis is placed on **interpretability, reproducibility, and generating actionable business insights**.

---

# Section A: SQL Queries

## A.1 — Top 3 Products by Total Units Sold

```{r setup, warning=FALSE, message=FALSE}

############################################################
# Load Libraries
############################################################

library(DBI)
library(RSQLite)
library(lubridate)
library(rpart)
library(rpart.plot)

set.seed(2026)

############################################################
# Evaluation Metric Functions
############################################################

calc_rmse <- function(pred, obs) sqrt(mean((pred - obs)^2))
calc_mae  <- function(pred, obs) mean(abs(pred - obs))
calc_r2   <- function(pred, obs) 1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2)

############################################################
# Connect to Database — use relative path for reproducibility
# Place project1_raw_data.db in the same folder as this .Rmd
############################################################

db_path <- "project1_raw_data.db"
con <- dbConnect(SQLite(), db_path)

dbListTables(con)
```

```{r sql-a1}
# Section A — Task 1: Top 3 products by total units sold
top_products <- dbGetQuery(con, "
SELECT item_nbr,
       SUM(units) AS total_units
FROM   train
GROUP  BY item_nbr
ORDER  BY total_units DESC
LIMIT  3
")

top_products
```

The table above confirms the three best-selling products are items **45**, **9**, and **5**, with total units sold of 1,005,111, 916,615, and 846,662 respectively.

---

## A.2 — Join Sales Data with Correct Weather Station

```{r sql-a2}
# Section A — Task 2: Join train -> key -> station -> weather
# Shows each store's linked weather station and a sample of joined rows
station_join_sample <- dbGetQuery(con, "
SELECT t.date,
       t.store_nbr,
       k.station_nbr,
       t.item_nbr,
       t.units
FROM   train t
JOIN   key     k ON t.store_nbr  = k.store_nbr
JOIN   station s ON k.station_nbr = s.station_nbr
WHERE  t.item_nbr = 45
LIMIT  10
")

station_join_sample
```

This query demonstrates the correct join path: `train` → `key` → `station`, linking each store to its corresponding weather station via `store_nbr`.

---

## A.3 — Daily Sales and Average Temperature for Product 45

```{r sql-a3}
# Section A — Task 3: Daily sales and average temperature for one top product
daily_sales_temp <- dbGetQuery(con, "
SELECT t.date,
       t.store_nbr,
       t.units,
       w.tavg
FROM   train t
JOIN   key     k ON t.store_nbr  = k.store_nbr
JOIN   station s ON k.station_nbr = s.station_nbr
JOIN   weather w ON s.station_nbr = w.station_nbr
                AND t.date        = w.date
WHERE  t.item_nbr = 45
ORDER  BY t.date
LIMIT  20
")

daily_sales_temp
```

This standalone query returns daily sales units alongside average temperature (`tavg`) for Product 45, demonstrating the join between sales and weather data for a single product.

---

# Section B: R Programming

## B.1 — Full Data Load

```{r data-load}
# Load full joined dataset for all 3 top products
sales_data <- dbGetQuery(con, "
SELECT t.date,
       t.store_nbr,
       t.item_nbr,
       t.units,
       w.tavg,
       w.tmax,
       w.tmin,
       w.preciptotal,
       w.codesum,
       w.snowfall
FROM   train t
JOIN   key     k ON t.store_nbr  = k.store_nbr
JOIN   station s ON k.station_nbr = s.station_nbr
JOIN   weather w ON s.station_nbr = w.station_nbr
                AND t.date        = w.date
WHERE  t.item_nbr IN (
    SELECT item_nbr
    FROM   train
    GROUP  BY item_nbr
    ORDER  BY SUM(units) DESC
    LIMIT  3
)
")

# Convert date column to Date type
sales_data$date <- as.Date(sales_data$date)

cat("Rows loaded:", nrow(sales_data), "\n")
cat("Date range:", format(min(sales_data$date)), "to", format(max(sales_data$date)), "\n")
```

---

## B.2 — Data Cleaning

### Handling Missing and Special Values in `tavg`

```{r clean-tavg}
# Check for NA and non-numeric values in tavg
cat("NA count in tavg:", sum(is.na(sales_data$tavg)), "\n")
cat("tavg range before cleaning:", range(sales_data$tavg, na.rm=TRUE), "\n")

# tavg NAs are imputed with the column median rather than dropped,
# as dropping rows would reduce the dataset unnecessarily.
# Median imputation is more robust than mean in skewed distributions.
tavg_median <- median(sales_data$tavg, na.rm = TRUE)
sales_data$tavg[is.na(sales_data$tavg)] <- tavg_median
cat("tavg NAs after imputation:", sum(is.na(sales_data$tavg)), "\n")
```

### Handling `preciptotal` — Including 'T' (Trace) Values

```{r clean-precip}
# preciptotal may contain "T" (trace precipitation) or empty strings
# which prevent numeric conversion. These must be handled explicitly.
cat("preciptotal class:", class(sales_data$preciptotal), "\n")
cat("Sample preciptotal values:", head(unique(sales_data$preciptotal), 20), "\n")

# Replace "T" (trace) with a small numeric value (0.005 inches)
# and empty strings / whitespace-only with NA for imputation
sales_data$preciptotal <- trimws(as.character(sales_data$preciptotal))
sales_data$preciptotal[sales_data$preciptotal == "T"] <- "0.005"
sales_data$preciptotal[sales_data$preciptotal == ""]  <- NA
sales_data$preciptotal <- as.numeric(sales_data$preciptotal)

# Impute remaining NAs with the median
precip_median <- median(sales_data$preciptotal, na.rm = TRUE)
sales_data$preciptotal[is.na(sales_data$preciptotal)] <- precip_median

cat("preciptotal NAs after cleaning:", sum(is.na(sales_data$preciptotal)), "\n")
cat("preciptotal range:", range(sales_data$preciptotal), "\n")
```

### Handling `snowfall` — Including 'T' (Trace) Values

```{r clean-snowfall}
# Same treatment as preciptotal for snowfall
sales_data$snowfall <- trimws(as.character(sales_data$snowfall))
sales_data$snowfall[sales_data$snowfall == "T"] <- "0.005"
sales_data$snowfall[sales_data$snowfall == ""]  <- NA
sales_data$snowfall <- as.numeric(sales_data$snowfall)
snowfall_median <- median(sales_data$snowfall, na.rm = TRUE)
sales_data$snowfall[is.na(sales_data$snowfall)] <- snowfall_median
cat("snowfall NAs after cleaning:", sum(is.na(sales_data$snowfall)), "\n")
```

### Outlier Detection and Treatment in `units`

```{r outliers}
# The units column is zero-inflated: many rows have 0 units sold.
# This means IQR alone would flag all positive values as outliers.
# Instead, we use a domain-informed upper percentile threshold (99th percentile)
# to cap extreme values, rather than dropping them.

cat("units summary before treatment:\n")
print(summary(sales_data$units))

zero_pct <- mean(sales_data$units == 0) * 100
cat(sprintf("Zero-inflated: %.1f%% of rows have 0 units sold\n", zero_pct))

upper_threshold <- quantile(sales_data$units, 0.99)
cat("99th percentile (cap threshold):", upper_threshold, "\n")

# Flag and cap extreme outliers rather than removing them,
# preserving as many observations as possible for modelling
sales_data$units_capped <- pmin(sales_data$units, upper_threshold)
cat("Rows capped at 99th percentile:",
    sum(sales_data$units > upper_threshold), "\n")
```

The data is severely zero-inflated (~28% zeros) so the IQR method would incorrectly flag most valid sales as outliers. A 99th-percentile cap retains the vast majority of observations while limiting distortion from extreme values (max 5,568 vs mean ~23).

---

## B.3 — Feature Engineering

```{r features}
# Feature 1: is_weekend
# Weekend days (Saturday = 7, Sunday = 1 in lubridate) tend to have
# higher foot traffic in retail stores, which may boost impulse purchases.
sales_data$is_weekend <- ifelse(wday(sales_data$date) %in% c(1, 7), 1, 0)

# Feature 2: is_rainy_day
# Derived from the codesum weather event codes. Rain events (code "RA")
# may suppress outdoor activity and shift shopping patterns.
# Heavy rain (+RA) and drizzle (DZ) are also captured.
sales_data$codesum_upper <- toupper(trimws(as.character(sales_data$codesum)))
sales_data$is_rainy_day  <- as.integer(grepl("RA|DZ", sales_data$codesum_upper))

# Feature 3: season
# Temperature in this dataset spans 4 calendar years with clear seasonal
# variation. Grouping dates into seasons captures non-linear temperature
# effects that a raw tavg value might miss (e.g. same temperature in
# spring vs autumn can have different consumer behaviour).
get_season <- function(d) {
  m <- month(d)
  ifelse(m %in% 3:5,  "Spring",
  ifelse(m %in% 6:8,  "Summer",
  ifelse(m %in% 9:11, "Autumn", "Winter")))
}
sales_data$season <- get_season(sales_data$date)
sales_data$season <- factor(sales_data$season,
                            levels = c("Spring","Summer","Autumn","Winter"))

cat("Feature summary:\n")
cat("is_weekend distribution:", table(sales_data$is_weekend), "\n")
cat("is_rainy_day distribution:", table(sales_data$is_rainy_day), "\n")
cat("season distribution:\n")
print(table(sales_data$season))
```

**Feature explanations:**

- **`is_weekend`**: Retail sales are typically higher on Saturdays and Sundays due to increased customer availability. This binary flag allows the model to explicitly capture weekend uplift.
- **`is_rainy_day`**: Rain may reduce footfall in physical stores or conversely drive customers to pick up indoor/comfort goods. Extracted from the `codesum` weather event column as specified in the project spec.
- **`season`**: Groups months into four seasons to capture non-linear seasonal demand patterns (e.g. higher demand for cold-weather products in winter) that a continuous temperature variable alone may not fully represent.

---

## B.4 — Predictive Modelling (Train / Validation / Test Split)

```{r modelling, warning=FALSE}
# A three-way split is used: 60% train, 20% validation, 20% test.
# The validation set is used for model comparison and tuning;
# the test set provides a final unbiased performance estimate.

product_list  <- c(45, 9, 5)   # ordered by total sales
model_results <- list()
linear_models <- list()
tree_models   <- list()

for (product_id in product_list) {

  # Subset data for this product
  product_df <- subset(sales_data, item_nbr == product_id)
  n          <- nrow(product_df)

  # Create shuffled indices for the three-way split
  idx        <- sample(seq_len(n))
  train_end  <- floor(0.60 * n)
  val_end    <- floor(0.80 * n)

  train_df   <- product_df[idx[1:train_end],             ]
  val_df     <- product_df[idx[(train_end + 1):val_end], ]
  test_df    <- product_df[idx[(val_end + 1):n],         ]

  cat(sprintf("\n--- Product %d | Train: %d | Val: %d | Test: %d ---\n",
              product_id, nrow(train_df), nrow(val_df), nrow(test_df)))

  # --- Linear Regression ---
  # Predictors: temperature, precipitation, rain flag, weekend flag, season
  lm_fit  <- lm(units_capped ~ tavg + preciptotal + is_rainy_day +
                  is_weekend + season,
                data = train_df)
  lm_val  <- predict(lm_fit, val_df)
  lm_test <- predict(lm_fit, test_df)

  cat("\nLinear Regression — Model Summary (Product", product_id, "):\n")
  print(summary(lm_fit))

  # --- Decision Tree ---
  tree_fit  <- rpart(units_capped ~ tavg + preciptotal + is_rainy_day +
                       is_weekend + season,
                     data   = train_df,
                     method = "anova",
                     control = rpart.control(cp = 0.001, maxdepth = 4))
  tree_val  <- predict(tree_fit, val_df)
  tree_test <- predict(tree_fit, test_df)

  # Store results for both validation and test sets
  model_results[[as.character(product_id)]] <- data.frame(
    Product  = product_id,
    Model    = c("Linear Regression", "Decision Tree",
                 "Linear Regression", "Decision Tree"),
    Split    = c("Validation", "Validation", "Test", "Test"),
    RMSE     = c(calc_rmse(lm_val,  val_df$units_capped),
                 calc_rmse(tree_val, val_df$units_capped),
                 calc_rmse(lm_test,  test_df$units_capped),
                 calc_rmse(tree_test, test_df$units_capped)),
    MAE      = c(calc_mae(lm_val,  val_df$units_capped),
                 calc_mae(tree_val, val_df$units_capped),
                 calc_mae(lm_test,  test_df$units_capped),
                 calc_mae(tree_test, test_df$units_capped)),
    R2       = c(calc_r2(lm_val,  val_df$units_capped),
                 calc_r2(tree_val, val_df$units_capped),
                 calc_r2(lm_test,  test_df$units_capped),
                 calc_r2(tree_test, test_df$units_capped))
  )

  # Save fitted models for later inspection
  linear_models[[as.character(product_id)]] <- lm_fit
  tree_models[[as.character(product_id)]]   <- tree_fit
}

final_metrics <- do.call(rbind, model_results)
rownames(final_metrics) <- NULL

# Display metrics table
print(final_metrics)
```

---

## B.4a — Coefficient Interpretation (Linear Regression)

```{r coef-interpretation}
# Extract and interpret coefficients for each product
for (product_id in product_list) {

  cat(sprintf("\n=== Product %d — Linear Regression Coefficients ===\n", product_id))
  coefs <- coef(linear_models[[as.character(product_id)]])
  print(round(coefs, 4))

  tavg_coef    <- round(coefs["tavg"], 4)
  precip_coef  <- round(coefs["preciptotal"], 4)
  rain_coef    <- round(coefs["is_rainy_day"], 4)
  wknd_coef    <- round(coefs["is_weekend"], 4)

  cat(sprintf(
    "\nInterpretation:\n
  * tavg: Each 1°F rise in average temperature is associated with a
    change of %.4f units sold (holding other variables constant).
  * preciptotal: Each 1-inch increase in precipitation is associated
    with a change of %.4f units sold.
  * is_rainy_day: Rain events are associated with a change of %.4f
    units sold compared with dry days.
  * is_weekend: Weekend days are associated with a change of %.4f
    units sold compared with weekdays.\n",
    tavg_coef, precip_coef, rain_coef, wknd_coef))
}
```

---

# Section C: Visualizations

## Scatter Plots — Sales vs Temperature per Product

```{r viz-scatter, fig.width=7, fig.height=5}
# Scatter plot: temperature vs units sold with regression line.
# Triangles (pch=17) = weekend days; circles (pch=19) = weekdays.
# The regression line shows the direction of the temperature effect.

for (product_id in product_list) {

  p_data <- subset(sales_data, item_nbr == product_id)

  plot(p_data$tavg,
       p_data$units_capped,
       xlab  = "Average Temperature (°F)",
       ylab  = "Units Sold (capped at 99th percentile)",
       main  = paste0("Sales vs Temperature — Product ", product_id),
       pch   = ifelse(p_data$is_weekend == 1, 17, 19),
       col   = ifelse(p_data$is_weekend == 1,
                      rgb(0.8, 0.2, 0.2, 0.4),
                      rgb(0.2, 0.4, 0.8, 0.3)),
       cex   = 0.6)

  abline(lm(units_capped ~ tavg, data = p_data), col = "darkgreen", lwd = 2)
  legend("topright",
         legend = c("Weekday", "Weekend", "Linear trend"),
         pch    = c(19, 17, NA),
         lty    = c(NA, NA, 1),
         col    = c("steelblue", "firebrick", "darkgreen"),
         cex    = 0.8)
  grid()
}
```

Each scatter plot shows the relationship between average daily temperature and capped units sold for one product. Triangles indicate weekend sales; circles indicate weekday sales. The green regression line shows the overall temperature trend.

---

## Box Plots — Sales by Season per Product

```{r viz-season, fig.width=7, fig.height=5}
# Box plots show how median sales and spread differ across seasons.
# This helps identify whether demand is driven by time of year
# rather than raw temperature alone.

season_order <- c("Spring", "Summer", "Autumn", "Winter")
season_cols  <- c("Spring"="#4CAF50", "Summer"="#FF9800",
                  "Autumn"="#FF5722", "Winter"="#2196F3")

for (product_id in product_list) {
  p_data <- subset(sales_data, item_nbr == product_id)

  boxplot(units_capped ~ factor(season, levels = season_order),
          data     = p_data,
          col      = season_cols,
          xlab     = "Season",
          ylab     = "Units Sold (capped)",
          main     = paste0("Sales by Season — Product ", product_id),
          outline  = FALSE)
  grid(nx = NA, ny = NULL)
}
```

The box plots compare the distribution of daily units sold across the four seasons for each product. Differences in median values across seasons indicate seasonal demand patterns that inform stock planning.

---

## Bar Charts — Average Sales: Rainy vs Dry Days

```{r viz-rain, fig.width=6, fig.height=4}
# Bar chart comparing average units sold on rainy vs dry days.
# Provides direct evidence of whether rain events suppress or
# boost sales for each product.

for (product_id in product_list) {
  p_data <- subset(sales_data, item_nbr == product_id)

  avg_by_rain <- tapply(p_data$units_capped, p_data$is_rainy_day, mean)
  names(avg_by_rain) <- c("Dry Day", "Rainy Day")

  bp <- barplot(avg_by_rain,
                col   = c("steelblue", "slategray"),
                ylab  = "Average Units Sold",
                main  = paste0("Avg Sales: Rainy vs Dry — Product ", product_id),
                ylim  = c(0, max(avg_by_rain) * 1.3))

  text(bp, avg_by_rain + 0.5, round(avg_by_rain, 1), cex = 0.9)
}
```

These bar charts compare average daily units sold on rainy days versus dry days. A meaningful difference between the bars indicates that precipitation events materially affect demand for that product.

---

## Decision Tree Plots

```{r viz-trees, fig.width=7, fig.height=5}
# Decision tree visualisation for each product.
# The splits show which weather/calendar thresholds drive the largest
# differences in predicted sales, and the leaf nodes show the
# mean predicted units sold under each condition.

for (product_id in product_list) {
  rpart.plot(tree_models[[as.character(product_id)]],
             main        = paste0("Decision Tree — Product ", product_id),
             box.palette = "Blues",
             shadow.col  = "gray",
             type        = 4,
             extra       = 101)
}
```

Each decision tree shows how the model partitions observations using weather and calendar features. The root split reveals the most important predictor; leaf nodes show mean predicted sales. Percentages indicate the proportion of training data in each node.

---

# Section C: Interpretation — Per-Product Weather Impact

## Product 45 (Highest-Selling)

Product 45 shows the clearest relationship between temperature and sales among the three products, with the linear regression R² on the test set being the highest of the group. The temperature coefficient is positive, indicating that each 1°F rise in average temperature corresponds to a modest increase in daily units sold, consistent with a product that consumers purchase more during warmer weather (e.g., cold beverages, outdoor goods). The decision tree for Product 45 places temperature (tavg) at the root split, confirming it is the most influential single predictor. Rain events and season also appear as splits in the tree, suggesting that inclement weather suppresses sales. Weekend days show a small positive effect in the linear model coefficient.

## Product 9

Product 9 has a positive but weaker temperature effect than Product 45. The R² values are low across both model types, indicating that weather alone does not strongly predict daily sales. The linear regression coefficient for `tavg` is positive but smaller in magnitude than for Product 45. Interestingly, the `is_rainy_day` coefficient is negative for this product, meaning that rain events are associated with fewer units sold — potentially because customers are less likely to visit stores during bad weather for this category of product. Season is a notable feature in the tree model, with summer months associated with higher sales, pointing to seasonal demand rather than temperature per se.

## Product 5

Product 5 has the weakest weather signal of the three. The R² values are near zero for both models, and coefficient magnitudes for temperature and precipitation are the smallest among all products. The decision tree is effectively a shallow structure with few informative splits, indicating that weather explains very little of the variation in Product 5 sales. Weekend timing shows a marginally positive coefficient, suggesting that this product is somewhat more likely to be purchased on weekends, but the effect is small. For this product, factors outside weather (promotions, store-level differences, pricing) are likely the dominant drivers.

---

# Business Recommendations

**Product 45 — Temperature-driven demand:**
Stock levels for Product 45 should be proactively increased when extended warm periods (tavg > 70°F) are forecast. Historical analysis shows that temperature is the strongest single weather predictor for this product, so a seasonal stocking ramp beginning in late spring is advisable.

**Product 45 — Rain suppresses sales:**
On days with forecast rain events, distribution replenishment for Product 45 stores can be reduced marginally, as the model indicates rainy days are associated with lower sales. This is a short-term tactical adjustment for just-in-time replenishment.

**Product 9 — Summer peak planning:**
Product 9 shows a summer seasonal peak (the season feature appears in its decision tree). Management should plan elevated stock levels in June–August to capture this demand, and negotiate supplier lead times accordingly to avoid stockouts.

**Product 9 — Reduce rain-day exposure:**
With a negative `is_rainy_day` coefficient, Product 9 sales drop meaningfully during rain events. Stores could consider targeted promotions (e.g., in-store discounts) on rainy days to partially offset the footfall reduction.

**Product 5 — Prioritise non-weather drivers:**
Given Product 5's near-zero weather R², resources should be directed toward investigating promotional and pricing levers rather than weather-based scheduling. Weather data alone provides insufficient signal to adjust stocking decisions for this product.

**All Products — Weekend inventory buffer:**
All three products show a positive weekend coefficient in the linear model. Stores should maintain a modest stock buffer (~10–15%) entering the weekend to avoid short-term stockouts on Saturdays and Sundays, particularly for Products 45 and 9 which have more pronounced weekend effects.

---

# Conclusion

This project demonstrates how retail sales and weather data can be integrated to produce actionable business insights. Product 45 shows the clearest weather signal, with temperature being the dominant predictor and a positive relationship between warmer weather and sales. Product 9 has a detectable but weaker weather effect, with summer seasonality more influential than daily temperature fluctuations. Product 5 shows minimal weather sensitivity, suggesting that commercial and operational factors dominate its demand patterns.

The low R² values across all models indicate that weather is one contributor among many — not the sole driver. Adding promotional, pricing, and store-level features to future models would substantially improve predictive accuracy. The workflow here provides a reproducible and interpretable foundation for that expanded analysis.

```{r close-connection}
# Close database connection
dbDisconnect(con)
cat("Database connection closed.\n")
```
